Algorithms for the Automated Correction of Vertical Drift in Eye Tracking Data

### 2020-10-06 manuscript/SI1.tex
Made pseudocode appendix into supplementary item

\documentclass[doc,biblatex]{apa7}
\DeclareLanguageMapping{american}{american-apa}
\addbibresource{refs.bib}
\usepackage{amssymb} % For producing checkmark character
\captionsetup[figure]{textfont=rm,font=small} % Override italic APA figure captions
\newcommand\possessivecite[1]{\citeauthor{#1}'s (\citeyear{#1})} % Possessive citation command

\title{Algorithms for the Automated Correction of Vertical Drift in Eye Tracking Data}
\shorttitle{Algorithms for the correction of vertical drift}

%%%%%%%%%%%%%%%%%%%%% PREPRINT VERSION %%%%%%%%%%%%%%%%%%%%%

\author{Jon W. Carr, Valentina N. Pescuma, Michele Furlan, Maria Ktori, Davide Crepaldi}

\affiliation{Cognitive Neuroscience, International School for Advanced Studies, Trieste, Italy}

\authornote{\textbf{Cite as:} Carr, J.\@ W., Pescuma, V.\@ N., Furlan, M., Ktori, M., \& Crepaldi, D.\@ (under review). Algorithms for the automated correction of vertical drift in eye tracking data. https://doi.org/10.31219/osf.io/jg3nc}

%%%%%%%%%%%%%%%%%%%% SUBMISSION VERSION %%%%%%%%%%%%%%%%%%%%

% \fiveauthors{Jon W. Carr}{Valentina N. Pescuma}{Michele Furlan}{Maria Ktori}{Davide Crepaldi}

% \fiveaffiliations{Cognitive Neuroscience, International School for Advanced Studies (SISSA)\\Via Bonomea 265, 34136 Trieste TS, Italy\\+39 338 360 3224, jcarr@sissa.it}{Cognitive Neuroscience, International School for Advanced Studies (SISSA)}{Cognitive Neuroscience, International School for Advanced Studies (SISSA)}{Cognitive Neuroscience, International School for Advanced Studies (SISSA)}{Cognitive Neuroscience, International School for Advanced Studies (SISSA)}

% \authornote{This work was funded by an ERC Starting Grant (no.\@ 679010, STATLEARN) and a FARE 2016 grant (no.\@ R164XBHNYR, CROWDLEARN), both awarded to \textbf{DC}. \textbf{VNP}, \textbf{MK}, and \textbf{DC} conceived the original idea behind this work. \textbf{JWC} and \textbf{VNP} explored the literature and selected the algorithms. \textbf{VNP} and \textbf{MF} developed the \texttt{split} algorithm, with input from \textbf{MK} and \textbf{DC}. \textbf{JWC} developed the \texttt{warp} algorithm, implemented all algorithms in a single framework, conducted the simulations, and performed the analyses. \textbf{JWC} and \textbf{VNP} performed the manual corrections and drafted the paper. \textbf{DC} revised the paper and coordinated the project.}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\abstract{A common problem in eye tracking research is vertical drift---the progressive displacement of fixation registrations on the vertical axis that results from a gradual loss of eye tracker calibration over time. This is particularly problematic in experiments that involve the reading of multiline passages, where it is critical that fixations on one line are not erroneously recorded on an adjacent line. Correction is often performed manually by the researcher, but this process is tedious, time-consuming, and prone to error and inconsistency. Various methods have previously been proposed for the automated, post-hoc correction of vertical drift in reading data, but these methods vary greatly, not just in terms of the algorithmic principles on which they are based, but also in terms of their availability, documentation, implementation languages, and so forth. Furthermore, these methods have largely been developed in isolation with little attempt to systematically evaluate them, meaning that drift correction techniques are moving forward blindly. We document nine major algorithms, including two that are novel to this paper, and evaluate them using both simulated and natural eye tracking data. Our results suggest that a method based on dynamic time warping offers great promise, but we also find that some algorithms are better suited than others to particular types of drift phenomena and reading behavior, allowing us to offer evidence-based advice on algorithm selection.}

\keywords{algorithms; dynamic time warping; eye tracking; reading; vertical drift}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Reading is a fundamental skill for navigating modern society and, as such, is subject to intense study in the cognitive and language sciences. Among the many tools that researchers use to investigate reading in the laboratory, eye tracking occupies a prominent position. Using this technique, participants' eye movements may be recorded as they read written material, providing a window into the relevant cognitive processes as they unfold. Technological advancements in eye tracking, particularly from the 1970s \parencite[see e.g.,][]{Rayner:1998}, have allowed scholars to collect increasingly accurate measures of eye movements during reading tasks, leading to great improvements in the investigation of the cognitive processes underlying reading and reading acquisition.

Many eye tracking studies involve the reading of single words or sentences. For example, researchers may embed target words into different sentence contexts and manipulate predictability \parencite[e.g.,][]{Rayner:2001}, display isolated words to gain insight into how a reader's eye moves when processing a word \parencite[e.g.,][]{Vitu:2004}, or reveal parts of words in a gaze-contingent fashion to investigate parafoveal processing \parencite[e.g.,][]{Schotter:2012}. Sentence reading experiments have also been essential in revealing the cognitive processes behind different levels of written language processing, from the width of the perceptual span \parencite[e.g.,][]{Rayner:1986,Blythe:2009} to the effects that word length and frequency have on eye movements \parencite[e.g.,][]{Joseph:2009,TiffinRichards:2015}, as well as the effects of syntactic \parencite[e.g.,][]{Frazier:1982,Pickering:1998} and lexical \parencite[e.g.,][]{Sereno:2006} ambiguity.

In our everyday experience, however, we often do not encounter sentences in isolation; a good part of our reading experience involves connected text that is distributed over multiple lines. Therefore, experiments based on paragraph reading also provide insight into the reading experience, while allowing us to address levels of processing that are simply not available when one reads a single sentence, such as the role of broader context or the integration of syntactic relations across sentence boundaries \parencite{Jarodzka:2017}. Indeed, studies of multiline reading have become more prevalent in recent years, with researchers using passage reading tasks to investigate, for example, the effect of text- and participant-level characteristics on eye movements \parencite{Kuperman:2018} or of contextual facilitation on developing readers' eye movements \parencite{TiffinRichards:2020}. Eye tracking databases of multiline reading, such as GECO \parencite{Cop:2017} and Provo \parencite{Luke:2018}, have also been made available.

A technical issue that arises from the particular circumstances of multiline reading is so-called ``vertical drift,'' which we define as the progressive displacement of fixation registrations on the vertical axis \textit{over time}. In other words, fixations may be recorded above or below the line of text that the participant was actually reading, and the amount of measurement error may fluctuate over time, making it nontrivial to eliminate. Fig.~\ref{fig01}A depicts a reading trial exhibiting vertical drift phenomena; in this case, fixations---especially those on the left hand side---are recorded around one line higher than where the reader was actually fixating, but they also tend to slope down to the right such that fixations on the right hand side seem to be better aligned.

Vertical drift can occur quite unpredictably, even following good quality calibration, and it is likely caused by subtle movements in head position or changes in pupil dilation, which can be difficult to control for, even in a laboratory setting. Such sources of measurement error are often exacerbated in the context of multiline reading because the reading of a passage of text necessarily takes longer than single words or sentences, during which time calibration may begin to degrade. In addition, there are less frequent opportunities to recalibrate the device, since calibration can only be performed between trials and not during the reading of a passage.

	\begin{figure}
	\makebox[\textwidth][c]{\includegraphics[]{figs/fig01_double_column.eps}}
	\vspace*{2pt}
	\caption{Example reading trials from an adult participant (left) and a child participant (right) taken from \textcite{Pescuma:0000}. Each dot represents a fixation and the size of the dots represents duration. The adult trial exhibits upward shift, especially in the lower left part of the passage. The child trial is extremely noisy and exhibits not just vertical drift issues but also many natural reading phenomena that will pose challenges to the algorithms.}
	\label{fig01}
	\end{figure}

Whatever the cause and however it manifests itself, vertical drift will ultimately have a negative impact on the analysis of eye tracking data because fixations will be mapped to words that were not actually being fixated at a given point in time (as we can see in Fig.~\ref{fig01}A). It is therefore incumbent on the researcher to recognize such issues when they occur and to take corrective measures. Specialized software packages, such as EyeLink Data Viewer (SR~Research, Toronto, Canada) or EyeDoctor (UMass Eyetracking Lab, Amherst, USA), provide the ability to manually move fixations, either individually or in small batches. However, manual realignment can be very time-consuming and is likely to be error-prone. In particular, the realignment process can be greatly complicated by other sources of noise or idiosyncratic reading behaviors. For example, Fig.~\ref{fig01}B depicts a reading trial by a child reader; in this case, not only is the data affected by drift issues, but there are also various natural reading behaviors, such as within- and between-line regressions, which add an additional layer of complexity to the task of realignment, not to mention the baseline level of noise and unusual features such as the arching sequence of fixations targeting line~4.

A number of methods have previously been developed to automate post-hoc vertical drift correction. \texttt{FixAlign}, an R package developed by \textcite{Cohen:2013}, is currently the most well-established method in the experimental psychology community, although other methods have recently been proposed by \textcite{Schroeder:0000} and \textcite{Spakov:2019}. In addition, there is a disparate body of work from several subfields of computer science, such as biometrics \parencite{Abdulin:2015}, educational technology \parencite{Hyrskykari:2006}, and user-interface design \parencite{Beymer:2005}, in which various ad-hoc algorithms have been reported \parencite[see also][]{LimaSanches:2015,Carl:2013,Mishra:2012,MartinezGomez:2012,Yamaya:2017}.

These reported methods can be difficult to evaluate and use because they vary widely in terms of their availability, design choices, implementation languages, usability, level of documentation, expected input data, and the extent to which they rely on project-specific heuristics or particular eye tracker hardware. Furthermore, these methods have largely been developed in isolation from each other, and there has been little attempt to systematically evaluate them, so drift correction software is moving forward blindly without an evidence base to support new directions. In this paper, we attempt to classify the reported methods into nine major approaches, which we formalize as nine simple algorithms that adopt a consistent design model. In other words, we do not attempt to evaluate existing software implementations; rather, we explore the spectrum of drift correction algorithms by isolating and evaluating the core principles on which previous methods have been based. Our goal is to provide a systematic comparison of these algorithms in order to guide researchers' choices about the most suitable methods and to lay a solid foundation for future drift correction software.

To be clear, the algorithms we consider in this paper are restricted to one specific type of problem. Firstly, we only consider algorithms intended for use on passages of text; other uses of eye tracking, such as visual search and scene perception, can also undergo drift correction, but the methods required are different \parencite[see e.g.,][]{Zhang:2011,Zhang:2014}. Secondly, we only consider the problem of post-hoc correction; vertical drift can also be corrected in real time, but this imposes a more restrictive set of constraints that are better handled by other types of algorithm \parencite[see e.g.,][]{Hyrskykari:2006,Palmer:2016}. Thirdly, we only consider fully automated algorithms that do not require human supervision; an alternative approach would make greater use of human insight.

The paper proceeds in four main sections. First, we outline the algorithms. Second, we test the algorithms on simulated fixation sequences afflicted with various types of measurement error. Third, we test the algorithms on an eye tracking dataset (two examples from which are presented in Fig.~\ref{fig01}). And finally, we discuss the major properties of the algorithms, provide guidance to researchers about their use, and suggest ways in which they can be improved further. All code and data required to reproduce the analyses reported in this paper, as well as Matlab/Octave, Python, and R implementations of the algorithms, are available from the public data archive associated with this paper: https://doi.org/10.17605/OSF.IO/7SRKG

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Algorithms}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

In this section, we describe nine algorithms for the automated, post-hoc correction of vertical drift. The reader may also wish to refer to the Appendix where the algorithms are presented in pseudocode alongside other technical details.

The \texttt{attach} algorithm is the simplest of the algorithms considered in this paper. The algorithm simply attaches each fixation to its closest line, as illustrated in Fig.~\ref{fig02}A. While this has the benefit of being extremely simple, it is generally not resilient to the kinds of drift phenomena described above. However, \texttt{attach} serves as a useful baseline algorithm, since it essentially corresponds to an eye tracking analysis in which no correction was performed---a standard analysis of eye tracking data would simply map fixations to the closest words or other areas of interest. We return to this point later in the paper.

	\begin{figure}
	\makebox[\textwidth][c]{\includegraphics[]{figs/fig02_double_column.eps}}
	\vspace*{2pt}
	\caption{Illustrations of the core principle of each algorithm. The colors represent which line each fixation was assigned to (black indicates line one, etc.). \texttt{attach} maps the fixation on ``sit'' to line 2, while all other algorithm correctly map it to line 1.}
	\label{fig02}
	\end{figure}

The \texttt{chain} algorithm is based closely on one of the methods implemented in the R package \texttt{popEye} \parencite{Schroeder:0000} and can be seen as an extension of \texttt{attach}. Fixations are first linked together into ``chains''---sequences of consecutive fixations that are within a specified \textit{x}~and \textit{y}~distance of each other (see Fig.~\ref{fig02}B for an example). Fixations within a chain are then attached to whichever line is closest to the mean of their \textit{y}~values. This procedure is similar to the methods reported by \textcite{Hyrskykari:2006} and \textcite{Mishra:2012}, so we consider these to be special cases of \texttt{chain}.

The \texttt{chain} algorithm generally provides better performance over \texttt{attach} by exploiting the sequence's order information, as illustrated in Fig.~\ref{fig02}B: The fixation on ``sit'' on line~1 is chained to two surrounding fixations and is correctly mapped to line~1; while \texttt{attach} assigns this fixation to line~2 to which it is technically closer. A disadvantage of the method, however, is that it is necessary to specify appropriate thresholds that determine when a new chain begins. If these thresholds are set too low, \texttt{chain} becomes equivalent to \texttt{attach}; if they are set too high, \texttt{chain} will group large numbers of fixations together and force them onto a single inappropriate line. By default, \texttt{popEye} sets the \textit{x}~threshold to 20 $\times$ the font height and the \textit{y}~threshold to 2 $\times$ the font height. It is not exactly clear how these defaults were chosen, but we would tentatively suggest that the \textit{x}~threshold should be set to approximately one long saccade length (we use 192~px), and the \textit{y}~threshold to around half a line height (we use 32~px).

The \texttt{cluster} algorithm is also based on one of the methods implemented in \texttt{popEye} \parencite{Schroeder:0000}. \texttt{cluster} applies \textit{k}-means clustering\footnote{We used the \texttt{KMeans} function from the Python library \texttt{scikit-learn} \parencite{scikit-learn}. There are many variations of \textit{k}-means clustering, which we have not systematically compared.} to the \textit{y}~values of all fixations in order to group the fixations into $m$ clusters, where $m$ is the number of lines in the passage. Once each fixation has been assigned to a cluster, clusters are mapped to lines based on the mean \textit{y}~values of their constituent fixations: The cluster with the smallest mean \textit{y}~value is assigned to line one and so forth.

Unlike \texttt{attach} and \texttt{chain}, \texttt{cluster} does not assign fixations to the closest line in absolute terms; instead, it operates on the principle that fixations with similar \textit{y}~values must belong to the same line regardless of how far away that line might be. As such, the algorithm generally handles drift issues quite well. However, \texttt{cluster} will often not perform well if there is even quite mild overlap between fixations from different lines. In addition, since \textit{k}-means clustering is not guaranteed to converge on the same set of clusters on every run, the \texttt{cluster} algorithm is nondeterministic and can be somewhat unpredictable across multiple runs on the same reading trial. This is an important consideration from the point of view of reproducible research output.

The \texttt{compare} algorithm is directly based on the method reported by \textcite{LimaSanches:2015} and is very similar to the more complex methods described by \textcite{Yamaya:2017}. The fixation sequence is first segmented into ``gaze lines'' by identifying the return sweeps---long saccades that move the eye from the end of one line to the start of the next. The algorithm considers any saccade that moves from right to left by more than some threshold value (we use 512~px) to be a return sweep. Gaze lines are then matched to text lines based on a measure of similarity between them. \textcite{LimaSanches:2015} considered three measures of similarity and found dynamic time warping \parencite{Vintsyuk:1968,Sakoe:1978} to be the best method. Similarly, \textcite{Yamaya:2017} use the closely related Needleman--Wunsch algorithm \parencite{Needleman:1970}. In the example in Fig.~\ref{fig02}D, we only depict the first gaze line (black dots), which is compared to each line of text; the best match is with the first text line, as reflected in its low cost (i.e., the sum of the mapping lines is minimized).

The gaze lines and text lines are only compared in terms of their \textit{x}~values under the assumption that the fixations in a gaze line should have a good horizontal alignment with the centers of the words in the corresponding text line. Relying only on the \textit{x}~values helps the algorithm overcome vertical drift issues, but it is also problematic because the lines of text in a passage tend to be horizontally similar to each other; each line tends to contain a similar number of words that are of a similar length, resulting in potential ambiguity about how gaze lines and text lines should be matched up. To alleviate this issue, both \textcite{LimaSanches:2015} and \textcite{Yamaya:2017} only compare the gaze line to a certain number of nearby text lines (we set this parameter to 3, which is effectively the closest line plus one line above and one line below).

The \texttt{merge} algorithm is closely based on the post-hoc correction method described by \textcite{Spakov:2019}. The algorithm begins by creating ``progressive sequences''---consecutive fixations that are sufficiently close together. This is similar to \texttt{chain}, except that the sequences are strictly progressive (they only move forward), so a regression will initiate a new progressive sequence. The original method uses several parameters to define what constitutes ``sufficiently close together,'' but here we simplify this down to a single parameter, the \texttt{y\_thresh} which determines how close the \textit{y}~values of two consecutive fixations must be to be considered part of the same progressive sequence (we use 32~px).

Once these sequences have been created, they are repeatedly merged into larger and larger sequences until the number of sequences is reduced to $m$---one for each line of text. On each iteration of the merge process, the algorithm fits a regression line to every possible pair of sequences (with the proviso that the two sequences must contain some minimum number of fixations). If the absolute gradient of the regression line or its error (root-mean-square deviation) is above a threshold (we use 0.1 and 20 respectively), the candidate merger is abandoned. The intuition here is that, if two sequences belong to the same text line, the regression line fit to their combined fixations will have a gradient close to 0 and low regression error. Of the candidate mergers that remain, the pair of sequences with the lowest error are merged and added to the pool of sequences, replacing the original two sequences and reducing their number by one. This process is repeated until no further mergers are possible.

The algorithm then enters the next ``phase'' of the process, in which the criteria are slightly relaxed, allowing more mergers to occur. These phases could in principle be defined by the user, but we follow the four-phase model reported by \textcite{Spakov:2019}, which effectively builds a set of heuristics into the algorithm. In Phase~1, the first and second sequences must each contain a minimum of 3~fixations to be considered for merging; in Phase~2, only the second sequence must contain a minimum of 3~fixations; in Phase~3, there is no minimum number of fixations; and in Phase~4, the gradient and regression error criteria are also entirely removed. Of course, as soon as the number of sequences is reduced to $m$ the algorithm exits the merge process, so not all four phases will necessarily be required. Finally, the set of $m$ sequences is matched to the set of text lines in positional order: The sequence with the smallest mean \textit{y}~value is mapped to line one and so forth.

A similar sounding method is reported by \textcite{Beymer:2005} whose technique is based on ``growing'' a gaze line by incrementally adding fixations until this results in a poor fit to a regression line, at which point a new gaze line is begun. However, the description of the method lacked sufficient detail for us to consider it further.

The \texttt{regress} algorithm, which is closely based on \possessivecite{Cohen:2013} R package \texttt{FixAlign}, treats the fixations as a cloud of unordered points and fits $m$ regression lines to this cloud. These regression lines are parameterized by a slope, vertical offset, and standard deviation, and the best parameters are obtained by minimizing\footnote{We used the \texttt{minimize} function from the Python library \texttt{SciPy} \parencite{Virtanen:2020}. We have not systematically compared the choice of optimizer settings.} an objective function that determines the overall fit of the lines through the fixations. The algorithm has six free parameters which are used to specify the lower and upper bounds of the slope, offset, and standard deviation. Here we directly adopt \texttt{FixAlign}'s defaults: $[-0.1, 0.1]$, $[-50, 50]$, and $[1, 20]$ respectively. Once the $m$ best-fitting regression lines are obtained (e.g., the dashed lines in Fig.~\ref{fig02}F), \texttt{regress} assigns each fixation to the highest-likelihood regression line, which in~turn is associated with a text line.

\texttt{regress} has some conceptual similarities with \texttt{merge} but differs in several important respects. Notably, \texttt{regress} takes a top-down approach, where \texttt{merge} is more bottom-up, and the regression lines that \texttt{regress} fits to the fixations cannot take independent values---it is assumed that all fixations are sloping in the same direction, with the same vertical offset, and with the same amount of within-line variance. In addition, unlike \texttt{merge}, \texttt{regress} does not utilize the order information; instead, like \texttt{cluster}, it views the fixations as a collection of unordered points.

The \texttt{segment} algorithm is a slight simplification of the method described by \textcite{Abdulin:2015}. The fixation sequence is first segmented into $m$ disjoint subsequences based on the $m-1$ most extreme backward saccades along the \textit{x}-axis (i.e., the saccades that are most likely to be return sweeps). These subsequences are then mapped to the lines of text chronologically, under the assumption that the lines of text will be read in order. \textcite{Abdulin:2015} do not state precisely how they identify the return sweeps, but it seems they potentially allow for more than $m$ subsequences to be identified, in which case, rereadings of a previous line, based on a threshold level of similarity, are discarded. The version of the algorithm considered here does not discard any fixations and instead always identifies exactly $m$ subsequences.

The advantage of this general approach, as emphasized by \textcite{Abdulin:2015}, is that the \textit{y}~values of the fixations are completely ignored, rendering any vertical drift entirely invisible to the algorithm. However, the approach does not allow for the possibility that the lines of text might be read out of order or that a line of text might be read more than once, which is not uncommon in normal reading behavior. Therefore, the great strength of \texttt{segment}---its identification of $m$ consecutive subsequences, permitting a chronological, as opposed to positional, mapping---is also its great weakness: If a large regression is mistakenly identified as a return sweep, this will lead to a catastrophic off-by-one error in subsequent line assignments.

As far as we know, the \texttt{split} algorithm takes an approach that is distinct from anything previously reported, although it is conceptually similar to \texttt{segment}. Like \texttt{segment}, the \texttt{split} algorithm works on the principle of splitting the fixation sequence into subsequences by first identifying the return sweeps. However, \texttt{split} is not restricted to finding exactly $m-1$ return sweeps; instead, it identifies the most likely set of return sweeps, however many that turns out to be. There are various ways of approaching this classification problem, but here we use \textit{k}-means clustering to partition the set of saccades into exactly two clusters. Since return sweeps are usually highly divergent from normal saccades, one of the two clusters will invariably contain the return sweeps, which can then be used to split the fixation sequence into subsequences. However, since this is not guaranteed to produce $m-1$ return sweeps (and therefore $m$ subsequences) an order-based mapping is not possible, so \texttt{split} must use absolute position: Subsequences are mapped to the closest text lines in absolute terms. \texttt{split} has the advantage of generally finding all true return sweeps, and even if it identifies some false positives, the resulting subsequences can still be mapped to the appropriate lines by absolute position. However, this also means the algorithm is less resilient to vertical drift issues.

The final algorithm we consider, \texttt{warp}, is also novel to this paper, although it is mostly a wrapper around a preexisting algorithm---dynamic time warping \parencite[DTW;][]{Vintsyuk:1968,Sakoe:1978}. DTW was used by the \texttt{compare} algorithm to provide a measure of dissimilarity between a gaze line and a text line. To our knowledge, however, there have been no previous reports of DTW being used for its primary purpose: finding the best alignment between two time sequences. This is somewhat surprising because DTW is the natural computational choice for tackling vertical drift. Two methods have previously been described that share some conceptual similarities. \textcite{Carl:2013} uses a basket of reading-related measures to place a cost on different paths through a lattice of fixation-to-character mappings and selects the path with minimal cost. \textcite{MartinezGomez:2012} convert the fixation sequence into a low resolution image and find the best transformation of this image to fit the overall shape of the passage. Both of these methods are quite convoluted, however, and we consider them to be special cases of \texttt{warp}, which is a direct application of the standard DTW algorithm to eye tracking data.

	\begin{figure}
	\makebox[\textwidth][c]{\includegraphics[]{figs/fig03_single_column.eps}}
	\vspace*{2pt}
	\caption{Illustration of the \texttt{warp} algorithm. The veridical fixation sequence is represented in blue, and the expected fixation sequence (the sequence of word centers) is represented in red. The dashed black lines show the DTW warping path---the optimal way to align the two sequences, such that the sum of the Euclidean distances between matched points (i.e., the sum of the dashed lines) is minimized.}
	\label{fig03}
	\end{figure}

In order to use DTW to realign the fixation sequence to the text, we first need to specify an expected fixation sequence. Since we expect the reader to traverse the passage from left to right and from top to bottom, we can use the series of word centers as the expected sequence, under the assumption that readers will target the centers of words \parencite{ORegan:1984}. Given the expected and veridical sequences as inputs, the DTW algorithm finds the optimal way to nonlinearly warp the sequences on the time dimension such that the overall Euclidean distance between matched points across the two sequences is minimized, while maintaining a monotonically increasing mapping.\footnote{Specifically, the first fixation must be mapped to (at least) the first word; the last fixation must be mapped to (at least) the last word; every other fixation must be mapped to at least one word; and, if fixation~$i$ is mapped to word~$j$, then fixation~$i+1$ must be mapped to word(s)~$\ge j$. And vice versa for the mapping from words to fixations.} In the ``warping path'' that results from this process, every fixation is mapped to one or more words and every word is mapped to one or more fixations (see Fig.~\ref{fig03} for an example). It is then simply a case of assigning each fixation to whichever line its mapped word(s) belong(s) to. In the unlikely event that the mapped words belong to different lines, the majority line wins or an arbitrary choice is made in the case of ties.

If the final fixation on line~$i$ were mapped to the first word on line~$i+1$, this would result in a large increase in the overall cost of the mapping, so line changes act as major clues about the best alignment. The upshot of this is that \texttt{warp} effectively segments the fixation sequence into exactly $m$ subsequences, which are mapped to the lines of text in chronological order. In this sense, \texttt{warp} behaves very much like \texttt{segment}. However, the additional benefit of \texttt{warp} is that it can simultaneously consider different possibilities about which saccades are the return sweeps, selecting only those that result in the best fit to the passage at a global level. Nevertheless, \texttt{warp} is ultimately limited by the veracity of the expected fixation sequence, which encodes one particular way of reading the passage---line by line from start to end. If the reader deviates from this assumption, \texttt{warp} can fail to correctly assign fixations to lines.

\subsection{Summary}

In this section we have described nine algorithms for aligning a fixation sequence to a multiline text, each of which takes a fundamentally different approach. A summary of the information utilized by the algorithms is provided in Table~\ref{table1}; each algorithm uses at~least one piece of information about the fixations and at~least one piece of information about the passage, and some also rely on additional parameters set by the user or built-in heuristics.

	\begin{table}
	\begin{center}
	\begin{threeparttable}
	\caption{Information Utilized by the Algorithms}
	\begin{tabular}{lcccccccc}
	\hline
	& \multicolumn{3}{c}{Fixation Info.} & \multicolumn{3}{c}{Passage Information} & \multicolumn{2}{c}{Other Information} \\
	Algorithm & \textit{X} & \textit{Y} & Order & No.\@ Lines & Line \textit{Y} & Word \textit{X} & Params. & Heuristics \\ \hline
	\texttt{attach}  & & \checkmark & & & \checkmark & & & \\
	\texttt{chain}   & \checkmark & \checkmark & \checkmark & & \checkmark & & 2 & \\
	\texttt{cluster} & & \checkmark & & \checkmark & & & & \\
	\texttt{compare} & \checkmark & \checkmark & \checkmark & & \checkmark & \checkmark & 2 & \\
	\texttt{merge}   & \checkmark & \checkmark & \checkmark & \checkmark & & & 3 & \checkmark \\
	\texttt{regress} & \checkmark & \checkmark & & \checkmark & \checkmark & & 6 & \\
	\texttt{segment} & \checkmark & & \checkmark & \checkmark & & & & \\
	\texttt{split}   & \checkmark & \checkmark & \checkmark & & \checkmark & & & \\
	\texttt{warp}    & \checkmark & \checkmark & \checkmark & & \checkmark & \checkmark & & \\ \hline
	\end{tabular}
	\label{table1}
	\end{threeparttable}
	\end{center}
	\end{table}

Broadly speaking, the algorithms proceed in three stages: grouping, assignment, and update. In the grouping stage, the fixations are, in~some~sense, placed into groups, the one exception being \texttt{attach} which has no grouping stage. The rational for how a group is formed varies by algorithm, but in general the algorithms can be categorized into those that are guaranteed to produce $m$ groups (i.e., one per text line; \texttt{cluster}, \texttt{merge}, \texttt{regress}, \texttt{segment}, and \texttt{warp}) and those that are not (\texttt{attach}, \texttt{chain}, \texttt{compare}, and \texttt{split}).

	\begin{table}
	\begin{center}
	\begin{threeparttable}
	\caption{Summary of the Grouping and Assignment Stages of each Algorithm}
	\begin{tabular}{l p{6.5cm} p{6.5cm}}
	\hline
	Algorithm        & Grouping Stage & Assignment Stage \\ \hline
	\texttt{attach}  & N/A & Assign fixations to closest text lines \\
	\texttt{chain}   & Chain consecutive fixations that are sufficiently close to each other & Assign chains to closest text lines \\
	\texttt{cluster} & Classify fixations into $m$ clusters based on their \textit{y}~values & Assign clusters to text lines in positional order \\
	\texttt{compare} & Split fixation sequence into subsequences based on saccades that are longer than a threshold & Assign subsequences to text lines by measuring horizontal similarity with the words in neighboring text lines \\
	\texttt{merge}   & Form a set of progressive sequences and then reduce the set to $m$ by repeatedly merging those that appear to be on the same line & Assign merged sequences to text lines in positional order \\
	\texttt{regress} & Find $m$ regression lines that best fit the fixations and group fixations according to best fit regression lines & Assign groups to text lines in positional order \\
	\texttt{segment} & Segment fixation sequence into $m$ subsequences based on $m-1$ most-likely return sweeps & Assign subsequences to text lines in chronological order \\
	\texttt{split}   & Split fixation sequence into subsequences based on best candidate return sweeps & Assign subsequences to closest text lines \\
	\texttt{warp}    & Map fixations to word centers by finding a monotonically increasing mapping with minimal cost, effectively resulting in $m$ subsequences & Assign fixations to the lines that their mapped words belong to, effectively assigning subsequences to text lines in chronological order \\ \hline
	\end{tabular}
	\label{table2}
	\end{threeparttable}
	\end{center}
	\end{table}

In the assignment stage, the previously identified groups are assigned to text lines. If the grouping stage is \textit{not} guaranteed to produce $m$ groups, then assignment must be based on absolute position (or similarity in the case of \texttt{compare}, although it still uses absolute position to select neighboring lines to compare to). If the grouping stage \textit{is} guaranteed to produce $m$ groups, then they can be assigned to text lines based on order; this generally allows for better handling of vertical drift because absolute position is ignored. In the case of \texttt{cluster}, \texttt{merge}, and \texttt{regress}, which produce unordered groups at the grouping stage, groups are matched to text lines based on the order in which they are positioned vertically (i.e., mean \textit{y}~value). In the case of \texttt{segment} and \texttt{warp}, the groups are assigned to text lines in chronological order, which is only possible because these two algorithms produce subsequences that inherit the order of the original fixation sequence. An overview of the grouping and assignment methods is provided in Table~\ref{table2} for quick reference.

Finally, in the update stage, the original fixation sequence is modified to reflect the line assignments identified in the previous stage. In the versions of the algorithms reported in this paper, we always use the same update approach: The \textit{y}~values of the fixations are adjusted to the \textit{y}~values of the assigned lines, while the \textit{x}~values and the order of the fixations are always left untouched. In principle, however, there are other ways of performing the update stage (e.g., retaining the original \textit{y}-axis variance or discarding ambiguous fixations).

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Performance on Simulated Data}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

We now test the ability of each algorithm to correctly recover the intended lines from simulated fixation sequences. These fixation sequences are simulated with particular characteristics, allowing us to understand how the algorithms respond to specific, isolated phenomena.

\subsection{Method}

In each simulation, we generate a passage of ``Lorem ipsum'' dummy text consisting of between 8 and 12 lines with up to 80 characters per line and 64~px of line spacing. We then generate a fixation sequence consisting of one fixation for every word in the passage: The \textit{x}~value of a fixation ($f_x$) is set randomly within the word; the \textit{y}~value of a fixation ($f_y$) is calculated according to:

	\begin{equation}
	f_y = \mathcal{N}(l_y, d_\mathrm{noise}) + f_x d_\mathrm{slope} + l_y d_\mathrm{shift},
	\label{equation1}
	\end{equation}

\noindent where $l_y$ is the vertical center point of the intended line---the \textit{y}~value that the reader is targeting. This models three types of distortion: noise, slope, and shift. Additionally, we simulate two types of regression that are characteristic of normal reading behavior but which can nevertheless disrupt algorithmic correction. Together, these five phenomena are illustrated in Fig.~\ref{fig04} and described below.

	\begin{figure}
	\makebox[\textwidth][c]{\includegraphics[]{figs/fig04_single_column.eps}}
	\vspace*{2pt}
	\caption{Example simulated fixation sequences under five phenomena considered in this paper. The algorithms must overcome these phenomena in order to correctly infer the intended line of each fixation.}
	\label{fig04}
	\end{figure}

\subsubsection{Noise Distortion}

>>>>

\documentclass[doc,biblatex]{apa7}
\DeclareLanguageMapping{american}{american-apa}

\title{Algorithms for the Automated Correction of Vertical Drift in Eye Tracking Data}
\shorttitle{Supplementary Item 1}

\author{Jon W. Carr, Valentina N. Pescuma,\\Michele Furlan, Maria Ktori, Davide Crepaldi}
\affiliation{Cognitive Neuroscience, International School for Advanced Studies, Trieste, Italy}

@@

\section{Supplementary Item 1}

\noindent
For reference, we present the algorithms here in a pseudocode that should be clear to programmers of any high-level scientific computing language. Matlab/Octave, Python, and R implementations may be found at https://doi.org/10.17605/OSF.IO/7SRKG. We have generally emphasized readability over optimization, and we make very minimal assumptions about the input and output. Most of the algorithms take two inputs: \texttt{fixation\_XY}, an array of size $n \times 2$ representing the \textit{xy} positions of $n$ fixations, and \texttt{line\_Y}, an array of length $m$ representing the \textit{y}~positions of the $m$ lines of text. Some algorithms take slightly different input or additional arguments as detailed below. All algorithms return a modified \texttt{fixation\_XY} as output, in which only the \textit{y}~values have been adjusted.
